Episodic memory: mental time travel or a quantum `memory wave' function?

### 2021-01-11 paper/main.tex
first attempt at addressing R1 concerns

Although reactivating context is often \textit{described} as a sort of mental time travel back to the one specific moment being remembered, there is some apparent inconsistency between this description and how this process is actually characterized mathematically or measured neurally.  In particular, both the theory and neural measures described in the preceding paragraphs characterize episodic recall (mathematically) as reflecting the reactivation of a weighted blend of thoughts from a \textit{range} of previously experienced times-- not to any one \textit{specific} time.  This is not simply a matter of imprecision in mental time travel (i.e., a reflection of the low effective resolution with which we can revisit specific moments from our past).  Rather, when we retrieve the thoughts associated with a range of times, it is conceptually more like we are simultaneously visiting \textit{many} of our prior experiences, analogous to a quantum wave function spreading its probability mass over space (see \textit{Defining the {\normalfont quantum memory wave function} metaphor and comparing it to the {\normalfont mental time travel} metaphor}). The above human and rodent studies suggest that the brain maintains parallel representations of ongoing experience, each drifting at a different timescale.  The notion of thinking of any one particular moment from the past is incompatible with this multiple timescales view, in that our thoughts and brain activity patterns reflect information from a range of moments, even when we are not specifically engaged in remembering.  If what we consider to be a ``moment'' or ``now'' is guided by our neural representations of our ongoing experiences, then this implies we are continually experiencing \textit{many} ``nows.''  If our thoughts are continually spread across a range of times, the notion of mentally revisiting any particular moment loses its meaning.

@@

The \textit{mental time travel} metaphor is defined by Tulving as ``the type of awareness experienced when one thinks back to a \textit{specific moment} in one's personal past and consciously recollects some prior episode or state as it was previously experienced''~\citep[Fig.~\ref{fig:qwave}B;][emphasis added]{WheeEtal97}.  Whereas mental time travel refers to mentally revisiting to one specific moment in addition to now (potentially with some temporal blurring), the quantum memory wave framework implies that episodic memory entails spreading our awareness over \textit{many} moments (Fig.~\ref{fig:qwave}C).  In this way, the key distinction between these framings is whether remembering entails simultaneously revisiting one single moment versus many moments.  The two frameworks converge when we engage in \textit{chronesthesia}~\citep{Tulv02b}-- i.e., when we are consciously aware of subjective time.   Analogous to how measuring the physical location of a particle collapses its quantum wave function at the moment of measurement, engaging in chronesthesia can collapse our quantum memory wave function such that our thoughts temporarily converge on the moment we are revisiting.  This wave function expands again when our mental state changes in the next moment.

@@

Although the specific mix of contents in each moment of the episode is nearly unique, participants' subjective experiences of the episode entail weaving together those temporally separated moments into a cohesive narrative.  This is reflected in how people verbally recount the episode later.  If we examine the temporal autocorrelation matrices of participants' recalls of the television episode (characterized using the same topic model used to capture the content of the episode; Figs.~\ref{fig:corrmats}B, C), we see a different structure than that of the episode itself.  Although individual participants' recalls also exhibit periods of content stability punctuated by rapid change (visible in the block diagonal structure of the matrices in Fig.~\ref{fig:corrmats}C), the off-diagonal correlations between the thematic content of temporally distant events are substantially stronger in participants' recalls than in the original episode.  Another way of characterizing this phenomenon is to look at the autocorrelation functions (describing the correlation between different moments' topic vectors as a function of their temporal distance) for the episode versus participants' recalls (Fig.~\ref{fig:reinstatement}A).  Whereas the autocorrelations in the original episode's content rapidly fall to 0, autocorrelations in the content of participants' recalls appear to asymptote around an average of 0.2.  These analyses indicate that the way participants \textit{recount} temporally separated events imposes additional similarity on those events, beyond what was present in the original events themselves.  One can then ask: is this simply a matter of imprecision in recall, either with respect to participants' recalls themselves, or with respect to the quality with which they are characterized via semantic feature vectors?  Or might there be a functional role of that additional imposed similarity structure?

@@

In addition to serving as a framework for quantifying the dynamic content of experiences, these trajectories may be visualized geometrically by projecting word embedding representations of the events onto lower dimensional spaces~\citep[e.g., ][]{HeusEtal18a}; for additional discussion and counterpoints see~\cite{JollChan18}.  \cite{HeusEtal18c} applied this approach to the same dataset reflected in Figures~\ref{fig:corrmats} and~\ref{fig:reinstatement}, revealing another key property of how we remember.  Although there are substantial individual differences in the specific words people use to describe their shared experience of watching the same television episode, and how much detail they recollected, the overall shapes of those recalled experience trajectories are highly similar across people.  This has potentially exciting implications for what it means to communicate memories to other people: to effectively communicate, we often need to convey the ``essence'' (gross overall shape) of our experience.  It also has implications for how we learn by analogy: perhaps events that share schemas~\citep[e.g., ][]{BaldEtal18}  unfold in similar ways (e.g., have similar shapes).  Learning which shapes go with which schemas could provide a scaffolding for rapid learning and generalization of new experiences, to the extent that those new experiences match a previously learned schema (shape).

In their paper, \cite{HeusEtal18c} use the geometric shapes of experience trajectories (through word embedding space) to functionally distinguish high-level essential details of the experience from low-level (non-essential) details.  When the trajectory exhibits a sharp change in direction, this indicates a rapid shift in the thematic content of the episode.  \cite{HeusEtal18c} used hidden Markov models to detect these rapid shifts and segment the trajectories into discrete \textit{events}.  The high-level (essential) details of the episode are reflected in the shape defined by the sequence of coordinates of its events.  These are captured by low spatial frequency properties of the trajectory's shape.  Low-level details are characterized by (typically smaller-scale) within-event geometric patterns, which are captured by the high spatial frequency properties of the trajectory's shape.  A key finding of that work was that people who watched the episode recalled high-level details reliably, and in a similar way across people, whereas people varied substantially in how they remembered low-level details.  This suggests that our memory systems may prioritize certain types of information (e.g., high-level details) about our experience over others (e.g., low-level details) as we encode our experiences into memories or when we recount our prior experiences.

@@

For real-world memory phenomena, and in many naturalistic memory paradigms, solving the matching problem is more complex.  People often experience real-world events (and scenes from movies or stories) as rich sensory experiences that derive deep meaning from surrounding moments, other prior experiences, emotions, goals, high-order understanding of the current situation, and so on.  Many of these aspects of our real-world experiences are not immediately verbalizable-- and even when we are able to describe some specific aspects of what we are experiencing, there are typically many possible ways we could have done so (e.g., varying in level of detail, specific wording, objectivity versus subjectivity, contextual details, etc.).  Further, because the way we experience each moment depends in part on our unique prior experiences, even ostensibly shared experiences can \textit{seem} different to each individual.  The way we recount a given experience is often a reflection of our own idiosyncratic experience of the original event, and is colored and distorted in an idiosyncratic way by our memory systems.  This distorted memory must then be verbalized or communicated, which is itself an idiosyncratic (and potentially lossy, in the information theoretic sense) process.  Solving the matching problem under these constraints poses a substantial theoretical and methodological challenge.  Specifically, matching requires building explicit models of the dynamic conceptual content of our experiences and memories, and then estimating when and how that experienced and remembered content aligns.  The approach featured in Figure~\ref{fig:trajectories} uses a word embedding space that is defined by a topic model~\citep{BleiEtal03}.  Each coordinate in this space reflects the conceptual content underlying a set of words, such that two different text passages that describe the same conceptual content (but using different words, levels of detail, etc.) are mapped onto nearby points in the word embedding space.  When these models are sufficiently expressive and robust to solve the matching problem by modeling our experiences and memories within the same word embedding space, we can see a clear geometric alignment between the trajectories that our experiences (Fig.~\ref{fig:trajectories}A) and memories (Fig.~\ref{fig:trajectories}B) take through word embedding space~\citep[also see][]{HeusEtal18c}.

\subsection*{Defining the geometries of ongoing thoughts and memories}
Defining geometric spaces that can reliably capture the conceptual contents of our thoughts and experiences has become a major focus of computational science.  One family of models that has shown particular promise over the past several decades is broadly referred to as \textit{word embedding models} (Fig.~\ref{fig:embedding-models}).  The fundamental goal of these models is to take as input a sequence of words (text) and produce as output one or more feature vectors whose coordinates reflect their conceptual (semantic) content.  For example, when two conceptually similar documents are passed through these models, the goal is to produce two feature vectors that are closer (e.g., in Euclidean distance) than the feature vectors for two dissimilar documents.

Early word embedding models such as \textit{Latent Semantic Analysis}~\citep[LSA;][]{DeerEtal90, LandDuma97} derived text embeddings using word co-occurence statistics.  The logic driving this approach posits that documents that contain similar mixtures of words are likely to be conceptually similar, and words that appear in similar documents are also likely to be conceptually similar.  To quantify these intuitions, \cite{DeerEtal90} first constructed \textit{word count matrices} (Fig.~\ref{fig:embedding-models}A) for several text corpora (each comprising many thousands of documents).  Each entry in these matrices denotes the number of time a given document (row) contained a given word (column).  They then used the Singular Value Decomposition (SVD) to decompose the number-of-documents $~\times~$ number-of-words word count matrix into the product of a number-of-documents $\times~K$ \textit{document embeddings matrix}, a $K~\times~K$ \textit{singular values} matrix, and a $K~\times$ number-of-unique-words \textit{word embeddings matrix}.  The rows of the document embeddings matrix reflect the similarities between documents, and the columns of the word embeddings matrix reflect similarities between words.

@@

\begin{figure}[tp]
\centering
\includegraphics[width=0.8\textwidth]{figs/word_embedding_models}
\caption{\textbf{Three approaches to text embedding.}  \textbf{A.~Matrix factorization models.} Per-document (or per-stimulus) word counts and/or frequencies are collected into a word count matrix (left).  To compute $K$-dimensional embeddings of the documents and words, the word count matrix is factorized (typically by computing its Singular Value Decomposition) into the product of a number-of-documents $\times~K$ \textit{document embeddings matrix}, a $K \times K$ \textit{singular values} matrix, and a $K~\times $ number-of-unique-words \textit{word embeddings matrix}.  \textbf{B.~Topic models.}  Fitting a topic model to a word count matrix (e.g., Panel A) reveals a set of latent \textit{topics}, or themes that pervade each document.  The topics are specified in a $K~\times$ number-of-words \textit{topics matrix}, and the per-document themes are specified in a number-of-documents $\times~K$ \textit{topic proportions matrix}. A given document's topic proportions reflect the specific mix of themes that are reflected in that document (bar graph on the left).  These proportions are discovered automatically by estimating the topic from which each word in the document is most probably drawn (highlighted words in \textit{Document text}).  \textbf{C.~Deep learning models.}  A wide variety of deep learning approaches to text embedding have been proposed.  Broadly, these approaches often combine a common set of elements, including \textit{transformers}~\citep[diagram adapted from][]{ViswEtal17}, \textit{deep averaging networks}~\citep[diagram adapted from][]{IyyeEtal15}, \textit{recurrent neural networks}~\citep[diagram adapted from][]{IyyeEtal15}, and \textit{deep autoencoders}~\citep[diagram adapted from][]{YousHame17}.}
\label{fig:embedding-models}
\end{figure}

\textit{Topic models}, such as \textit{Latent Dirichlet Allocation}~\citep[LDA;][]{BleiEtal03} represent an attempt to characterize the latent \textit{topics} or themes that documents comprise (Fig.~\ref{fig:embedding-models}B).  Whereas matrix factorization approaches represent semantic meanings using low-dimensional approximations of word co-occurrences or free association frequencies, topic models leverage these properties of documents (or behaviors) to explain \textit{why} words co-occur.  LDA defines a topic as a distribution of weights over words in the vocabulary.  For example, if one takes the \textit{vocabulary} to be the top 10,000 most commonly used words in English (excluding stop words that are devoid of explicit semantic meaning such as ``and,'' ``or,'' ``the,'' ``but,'' etc.), then each topic comprises a vector of 10,000 non-negative weights that sums to one.  As shown in the cartoon on the left of the Panel (where the four top-weighted words from each of five hypothetical topics are displayed), a fruit-related topic (shown in purple) might weight heavily on words like ``apple,'' ``banana,'' ``pear,'' and ``orange,'' whereas words like ``cat,'' ``dog,'' ``dolphin,'' and ``lobster'' might receive relatively small weights.  Given a set of $K$ topic vectors, topic models compute the blend of topics most probably reflected in a given document-- i.e., its \textit{topic proportions}.  These topic proportion estimates are made by examining how different topics weight each of the words in the document, and by assuming that documents tend to reflect few topics (rather than many topics).  Given only a word count matrix as input (just as in LSA), topic models automatically discover a $K~\times~$ number-of-words \textit{topics matrix}, which describes how each topic (row) weights each word in the vocabulary (column), and a number-of-documents $~\times~K$ \textit{topic proportions matrix}, which describes each document's topic proportions.  The rows of the topic proportions matrix may be treated as low-dimensional embeddings of the set of documents.

A key advantage of topic models over matrix factorization models is their ability to explain potentially ambiguous word usage.  For example, consider the use of the word ``bat'' in the following illustrative passage:
\begin{quote}
  \textit{``As the slavering mutant vampire \textbf{bat} swooped low to attack, its eyes ablaze, the explorer grunted and took a mighty swing of her trusty aluminum baseball \textbf{bat}.  The powerful strike connected with a echoing metallic \textit{clink}, stunning the wretched creature before it scrambled about on the cave floor and flew away.  She didn't even \textbf{bat} an eye-- as the Dark Wizard Chiroptera continued to draw strength from the mystic crystals, such things had become increasingly common of late, nearly to the point of mundanity...''}
  \end{quote}
At different ponts in the passage, ``bat'' refers to an animal, a piece of sports equipment, and an eye-related verb.  Because each of these uses of the word appear in the same (hypothetical) document, models driven solely by word co-occurrence statistics would blend these different meanings such that the embedding of ``bat'' was somewhat similar to other animals \textit{and} other pieces of sports equipment \textit{and} other eye-related verbs.  By contrast, treating each document as reflecting a blend of topics allows these uses to be disambiguated.  If other documents that mention the word ``bat'' describe vampires, swooping, or other mammals, but \textit{not} sports equipment or eye-related verbs, then a topic model could explain this pattern by inferring that multiple topics place substantial weight on the word ``bat.''  Another strength of topic models is their interpretability.  Because the models directly output each document's topic proportions along with each topic's weights on all words in the vocabulary, the general themes that each topic reflects are often intuitive and the per-document topic proportions may be directly related to those themes.

@@

@@

Many of these differences between physical and mental navigation also apply to other mental domains, including how we \textit{conceptualize} space~\citep[e.g., when we trace out a potential route in our mind;][]{GautvanW16, GautEtal18, ArzyScha19}.  Just as we mentally spread ourselves across a distribution of times when we remember, we can perform similar mental operations when we conceptualize and retrieve other types of information.  Further, these distributions may be estimated using existing methods.  For example, consider how we use pattern classifiers~\citep{NormEtal06b} to connect neural activity patterns to specific semantic~\citep{PolyEtal05a, MitcEtal08a, MannEtal12}, spatial~\citep{MillEtal13}, affective~\citep{ChanEtal18}, and social~\citep{MeyeEtal18} content being mentally manipulated.  A prevailing approach is to label specific brain activity patterns as each reflecting \textit{one particular} concept or mental state.  Variability in neural activity over repeated experiences or rememberings can result in classifiers spreading the probability masses of their predictions over multiple possible mental states, including non-contiguous mental states.  Traditional approaches then ``clean up'' this distribution over possible mental states by taking its average (expected), maximum, or modal value.  However, following the ``quantum wave function'' logic described in this manuscript, these multi-peaked receptive fields may provide a mechanism for associating non-contiguous representations.  What are some implications of this view for non-memory domains?

@@

The notion that variability and ``uncertainty'' can be informative also has implications for how our brains might efficiently represent information.  Recent work suggests that adversarial networks trained to learn a generative model of natural images~\citep[e.g., ][]{GatyEtal16, IsolEtal17} may be used to refine estimates of decoded visual stimuli~\citep{StYvNase18}. Related manifold learning approaches that ``snap'' decoded estimates to the nearest surface location on a learned manifold may be used in a similar way to decode sequences of movie frames~\citep{HeusEtal18a}.  Taken together, these findings suggest that not all \textit{possible} images (and by extension, all possible stimuli) are equally likely.  Rather, the set of stimuli likely to be encountered live within subspaces or on manifolds in the relevant representational space.  The geometries of these manifolds provides insights into the statistical structure of the stimuli.  They also provide potential insights into how the brain represents those stimuli, for example to the extent that the geometries of these manifolds may be leveraged to improve brain decoding.  The shapes of these manifolds can potentially provide insights into how our thoughts are distributed over representational spaces.

>>>>

Although reactivating context is often \textit{described} as a sort of mental time travel back to the one specific moment being remembered, there is some apparent inconsistency between this description and how this process is actually characterized mathematically or measured neurally.  In particular, both the theory and neural measures described in the preceding paragraphs characterize episodic recall (mathematically) as reflecting the reactivation of a weighted blend of thoughts from a \textit{range} of previously experienced times-- not to any one \textit{specific} time.  This is not simply a matter of imprecision in mental time travel (i.e., a reflection of the low effective resolution with which we can revisit specific moments from our past).  Rather, when we retrieve the thoughts associated with a range of times, it is conceptually more like we are simultaneously visiting \textit{many} of our prior experiences, analogous to a quantum wave function spreading its probability mass over space (see \textit{Defining the {\normalfont quantum memory wave function} metaphor and comparing it to the {\normalfont mental time travel} metaphor}). The above human and rodent studies suggest that the brain maintains parallel representations of ongoing experience, each reflecting information at a different timescale.  The notion of thinking of any one particular moment from the past is incompatible with this multiple timescales view, in that our thoughts and brain activity patterns reflect information from a range of moments, even when we are not specifically engaged in remembering.  If what we consider to be a ``moment'' or ``now'' is guided by our neural representations of our ongoing experiences, then this implies we are continually experiencing \textit{many} ``nows.''  If our thoughts are continually spread across a range of times, the notion of mentally revisiting any particular moment loses its meaning.

@@

The \textit{mental time travel} metaphor is defined by Tulving as ``the type of awareness experienced when one thinks back to a \textit{specific moment} in one's personal past and consciously recollects some prior episode or state as it was previously experienced''~\citep[Fig.~\ref{fig:qwave}B;][emphasis added]{WheeEtal97}.  Whereas mental time travel refers to mentally revisiting one specific moment in addition to now (potentially with some temporal blurring), the quantum memory wave framework implies that episodic memory entails spreading our awareness over \textit{many} moments (Fig.~\ref{fig:qwave}C).  In this way, the key distinction between these framings is whether remembering entails simultaneously revisiting one single moment versus many moments.  The two frameworks converge when we engage in \textit{chronesthesia}~\citep{Tulv02b}-- i.e., when we are consciously aware of subjective time.   Analogous to how measuring the physical location of a particle collapses its quantum wave function at the moment of measurement, engaging in chronesthesia can collapse our quantum memory wave function such that our thoughts temporarily converge on the moment we are revisiting.  This wave function expands again when our mental state changes in the next moment.

@@

Although the specific mix of contents in each moment of the episode is nearly unique, participants' subjective experiences of the episode entail weaving together those temporally separated moments into a cohesive narrative.  This is reflected in how people verbally recount the episode later.  If we examine the temporal autocorrelation matrices of participants' recalls of the television episode (characterized using the same topic model used to capture the content of the episode; Figs.~\ref{fig:corrmats}B, C), we see a different structure than that of the episode itself.  Although individual participants' recalls also exhibit periods of content stability punctuated by rapid change (visible in the blocky structure of the matrices in Fig.~\ref{fig:corrmats}C), the off-diagonal correlations between the thematic content of temporally distant events are substantially stronger in participants' recalls than in the original episode.  Another way of characterizing this phenomenon is to look at the autocorrelation functions (describing the correlation between different moments' topic vectors as a function of their temporal distance) for the episode versus participants' recalls (Fig.~\ref{fig:reinstatement}A).  Whereas the autocorrelations in the original episode's content rapidly fall to 0, autocorrelations in the content of participants' recalls appear to asymptote around an average of 0.2.  These analyses indicate that the way participants \textit{recount} temporally separated events imposes additional similarity on those events, beyond what was present in the original events themselves.  One can then ask: is this simply a matter of imprecision in recall, either with respect to participants' recalls themselves, or with respect to the quality with which they are characterized via semantic feature vectors?  Or might there be a functional role of that additional imposed similarity structure?

@@

In addition to serving as a framework for quantifying the dynamic content of experiences, these trajectories may be visualized by projecting word embedding representations of the events onto lower dimensional spaces~\citep[e.g., ][]{HeusEtal18a}; for additional discussion and counterpoints see~\cite{JollChan18}.  \cite{HeusEtal18c} applied this approach to the same dataset reflected in Figures~\ref{fig:corrmats} and~\ref{fig:reinstatement}, revealing another key property of how we remember.  Although there are substantial individual differences in the specific words people use to describe their shared experience of watching the same television episode, and how much detail they recollected, the overall shapes of those recalled experience trajectories are highly similar across people.  This has potentially exciting implications for what it means to communicate memories to other people: to effectively communicate, we often need to convey the ``essence'' (gross overall shape) of our experience.  It also has implications for how we learn by analogy: perhaps events that share schemas~\citep[e.g., ][]{BaldEtal18}  unfold in similar ways (e.g., have similar shapes).  Learning which shapes go with which schemas could provide a scaffolding for rapid learning and generalization of new experiences, to the extent that those new experiences match a previously learned schema (shape).

In their paper, \cite{HeusEtal18c} use the geometric shapes of experience trajectories (through word embedding space) to functionally distinguish high-level essential details of the experience from low-level (non-essential) details.  When the trajectory exhibits a sharp change in direction, this indicates a rapid shift in the thematic content of the episode.  \cite{HeusEtal18c} used hidden Markov models to detect these rapid shifts and segment the trajectories into discrete \textit{events}.  The high-level (essential) details of the episode are reflected in the shape defined by the sequence of coordinates of its events.  These are captured by low spatial frequency properties of the trajectory's shape.  Low-level details are characterized by (typically smaller-scale) within-event geometric patterns, which are captured by the high spatial frequency properties of the trajectory's shape.  A key finding of that work was that people who watched the episode recalled high-level details reliably, and in a similar way across people, whereas people varied substantially in how they remembered low-level details.  This suggests that our memory systems may prioritize certain types of information (e.g., high-level details) about our experiences over others (e.g., low-level details) as we encode our experiences into memories or when we recount our prior experiences.

@@

For real-world memory phenomena, and in many naturalistic memory paradigms, solving the matching problem is more complex.  People often experience real-world events (and scenes from movies or stories) as rich sensory experiences that derive deep meaning from surrounding moments, other prior experiences, emotions, goals, high-order understanding of the current situation, and so on.  Many of these aspects of our real-world experiences are not immediately verbalizable-- and even when we are able to describe some specific aspects of what we are experiencing, there are typically many possible ways we could have done so (e.g., varying in level of detail, specific wording, objectivity versus subjectivity, contextual details, etc.).  Further, because the way we experience each moment depends in part on our unique prior experiences, even ostensibly shared experiences can \textit{seem} different to each individual.  The way we recount a given experience is often a reflection of our own idiosyncratic experience of the original event, and is colored and distorted in an idiosyncratic way by our memory systems.  This distorted memory must then be verbalized or communicated, which is itself an idiosyncratic (and potentially lossy, in the information theoretic sense) process.  Solving the matching problem under these constraints poses a substantial theoretical and methodological challenge.  Specifically, matching requires building explicit models of the dynamic conceptual content of our experiences and memories, and then estimating when and how that experienced and remembered content aligns.  The approach featured in Figure~\ref{fig:trajectories} uses a word embedding space that is defined by a topic model~\citep{BleiEtal03}.  Each coordinate in this space reflects the conceptual content underlying a set of words, such that two different text passages that describe the same conceptual content (but using different words, levels of detail, etc.) are mapped onto nearby points in the word embedding space.  When these models are sufficiently expressive and robust to solve the matching problem, by modeling our experiences and memories within the same word embedding space, we can see a clear geometric alignment between the trajectories that our experiences (Fig.~\ref{fig:trajectories}A) and memories (Fig.~\ref{fig:trajectories}B) take through word embedding space~\citep[also see][]{HeusEtal18c}.

\subsection*{Defining the geometries of ongoing thoughts and memories}
Defining geometric spaces that can reliably capture the conceptual contents of our thoughts and experiences has become a major focus of computational science.  One family of models that has shown particular promise over the past several decades is broadly referred to as \textit{text embedding models} (Fig.~\ref{fig:embedding-models}).  The fundamental goal of these models is to take as input a sequence of words (text) and produce as output one or more feature vectors whose coordinates reflect their conceptual (semantic) content.  For example, when two conceptually similar documents are passed through these models, the goal is to produce two feature vectors that are closer (e.g., in Euclidean distance) than the feature vectors for two dissimilar documents.

When text embedding models succeed at capturing semantic overlap between different words or text passages, they can provide an elegant means of solving the matching problem.  For example, when two or more individuals describe a common experience, embedding the text of those descriptions in a high dimensional feature space abstracts away the specific word choices and grammatical choices in favor of the descriptions' underlying conceptual content.  Further, geometric differences in those embeddings (e.g., the Euclidean distances between the corresponding vectors) can capture subtle differences in conceptual content that might not have been readily apparent in the original text.  Of course, there are also limits on how accurately specific word embedding models characterize different conceptual aspects of a given text passage, and on how informative or meaningful those differences are from a psychological perspective.  The field of text embedding has advanced rapidly, particularly in recent years, enabling more accurate and informative approaches to characterizing the conceptual content of arbitrary text passages.

Early embedding models such as \textit{Latent Semantic Analysis}~\citep[LSA;][]{DeerEtal90, LandDuma97} derived text embeddings using word co-occurence statistics.  The logic driving this approach posits that documents that contain similar mixtures of words are likely to be conceptually similar, and words that appear in similar documents are also likely to be conceptually similar.  To quantify these intuitions, \cite{DeerEtal90} first constructed \textit{word count matrices} (Fig.~\ref{fig:embedding-models}A) for several text corpora (each comprising many thousands of documents).  Each entry in these matrices denotes the number of times a given document (row) contained a given word (column).  They then used the Singular Value Decomposition (SVD) to decompose the number-of-documents $~\times~$ number-of-words word count matrix into the product of a number-of-documents $\times~K$ \textit{document embeddings matrix}, a $K~\times~K$ \textit{singular values matrix}, and a $K~\times$ number-of-unique-words \textit{word embeddings matrix}.  The rows of the document embeddings matrix reflect the similarities between documents, and the columns of the word embeddings matrix reflect similarities between words.

@@

\begin{figure}[tp]
\centering
\includegraphics[width=0.8\textwidth]{figs/word_embedding_models}
\caption{\textbf{Three approaches to text embedding.}  \textbf{A.~Matrix factorization models.} Per-document (or per-stimulus) word counts and/or frequencies are collected into a word count matrix (left).  To compute $K$-dimensional embeddings of the documents and words, the word count matrix is factorized (typically by computing its Singular Value Decomposition) into the product of a number-of-documents $\times~K$ \textit{document embeddings matrix}, a $K \times K$ \textit{singular values matrix}, and a $K~\times $ number-of-unique-words \textit{word embeddings matrix}.  \textbf{B.~Topic models.}  Fitting a topic model to a word count matrix (e.g., Panel A) reveals a set of latent \textit{topics}, or themes that pervade each document.  The topics are specified in a $K~\times$ number-of-words \textit{topics matrix}, and the per-document themes are specified in a number-of-documents $\times~K$ \textit{topic proportions matrix}. A given document's topic proportions reflect the specific mix of themes that are reflected in that document (bar graph on the left).  These proportions are discovered automatically by estimating the topic from which each word in the document is most probably drawn (highlighted words in \textit{Document text}).  \textbf{C.~Deep learning models.}  A wide variety of deep learning approaches to text embedding have been proposed.  Broadly, these approaches often combine a common set of elements, including \textit{transformers}~\citep[diagram adapted from][]{ViswEtal17}, \textit{deep averaging networks}~\citep[diagram adapted from][]{IyyeEtal15}, \textit{recurrent neural networks}~\citep[diagram adapted from][]{IyyeEtal15}, and \textit{deep autoencoders}~\citep[diagram adapted from][]{YousHame17}.}
\label{fig:embedding-models}
\end{figure}

\textit{Topic models}, such as \textit{Latent Dirichlet Allocation}~\citep[LDA;][]{BleiEtal03} represent an attempt to characterize the latent \textit{topics} or themes that documents comprise (Fig.~\ref{fig:embedding-models}B).  Whereas matrix factorization approaches represent semantic meanings using low-dimensional approximations of word co-occurrences or free association frequencies, topic models leverage these properties of documents (or behaviors) to explain \textit{why} words co-occur.  LDA defines a topic as a distribution of weights over words in the vocabulary.  For example, if one takes the \textit{vocabulary} to be the top 10,000 most commonly used words in English (excluding stop words that are devoid of explicit semantic meaning such as ``and,'' ``or,'' ``the,'' ``but,'' etc.), then each topic comprises a vector of 10,000 non-negative weights that sum to one.  As shown in the cartoon on the left of the Panel (where the four top-weighted words from each of five hypothetical topics are displayed), a fruit-related topic (shown in purple) might weight heavily on words like ``apple,'' ``banana,'' ``pear,'' and ``orange,'' whereas words like ``cat,'' ``dog,'' ``dolphin,'' and ``lobster'' might receive relatively small weights.  Given a set of $K$ topic vectors, topic models compute the blend of topics most probably reflected in a given document-- i.e., its \textit{topic proportions}.  These topic proportion estimates are made by examining how different topics weight each of the words in the document, and by assuming that documents tend to reflect few topics (rather than many topics).  Given only a word count matrix as input (just as in LSA), topic models automatically discover a $K~\times~$ number-of-words \textit{topics matrix}, which describes how each topic (row) weights each word in the vocabulary (column), and a number-of-documents $~\times~K$ \textit{topic proportions matrix}, which describes each document's topic proportions.  The rows of the topic proportions matrix may be treated as low-dimensional embeddings of the set of documents.

A key advantage of topic models over matrix factorization models is their ability to explain potentially ambiguous word usage.  For example, consider the use of the word ``bat'' in the following illustrative passage:
\begin{quote}
  \textit{``As the slavering mutant vampire \textbf{bat} swooped low to attack, its eyes ablaze, the explorer grunted and took a mighty swing of her trusty aluminum baseball \textbf{bat}.  The powerful strike connected with an echoing metallic \textit{clink}, stunning the wretched creature before it scrambled about on the cave floor and flew away.  She didn't even \textbf{bat} an eye-- as the Dark Wizard Chiroptera continued to draw strength from the mystic crystals, such things had become increasingly common of late, nearly to the point of mundanity...''}
  \end{quote}
At different ponts in the passage, ``bat'' refers to an animal, a piece of sports equipment, and an eye-related verb.  Because each of these uses of the word appears in the same (hypothetical) document, models driven solely by word co-occurrence statistics would blend these different meanings such that the embedding of ``bat'' was somewhat similar to other animals \textit{and} other pieces of sports equipment \textit{and} other eye-related verbs.  By contrast, treating each document as reflecting a blend of topics allows these uses to be disambiguated.  If other documents that mention the word ``bat'' describe vampires, swooping, or other mammals, but \textit{not} sports equipment or eye-related verbs, then a topic model could explain this pattern by inferring that multiple topics place substantial weight on the word ``bat.''  Another strength of topic models is their interpretability.  Because the models directly output each document's topic proportions along with each topic's weights on all words in the vocabulary, the general themes that each topic reflects are often intuitive and the per-document topic proportions may be directly related to those themes.

@@

When the ``surface features'' of our experiences repeat at non-contiguous timepoints, or when one moment overlaps conceptually with others, this overlap can be characterized by text embedding models like those described above.  In turn, this can help explain non-contiguous ``bumps'' in the quantum memory wave function (e.g., Figs.~\ref{fig:qwave}C, \ref{fig:reinstatement}B) and off-diagonal peaks in semantic temporal correlation matrices (e.g., Fig.~\ref{fig:corrmats}).

@@

Many of these differences between physical and mental navigation also apply to other mental domains, including how we \textit{conceptualize} space~\citep[e.g., when we trace out a potential route in our mind;][]{GautvanW16, GautEtal18, ArzyScha19}.  Just as we mentally spread ourselves across a distribution of times when we remember, we can perform similar mental operations when we conceptualize and retrieve other types of information.  Further, these distributions may be estimated using existing methods.  For example, consider how we use pattern classifiers~\citep{NormEtal06b} to connect neural activity patterns to specific semantic~\citep{PolyEtal05a, MitcEtal08a, MannEtal12}, spatial~\citep{MillEtal13}, affective~\citep{ChanEtal18}, and social~\citep{MeyeEtal18} content being mentally manipulated.  A prevailing approach is to label specific brain activity patterns as each reflecting \textit{one particular} concept or mental state.  Variability in neural activity over repeated experiences or rememberings can result in classifiers spreading the probability masses of their predictions over multiple possible mental states, including non-contiguous mental states.  Traditional approaches then ``clean up'' this distribution over possible mental states by taking its average (expected), maximum, or modal value.  However, following the ``quantum wave function'' logic described in this manuscript, these multi-peaked response patterns may provide a mechanism for associating non-contiguous representations.  What are some implications of this view for non-memory domains?

@@

The notion that variability and ``uncertainty'' can be informative also has implications for how our brains might efficiently represent information.  Recent work suggests that adversarial networks trained to learn a generative model of natural images~\citep[e.g., ][]{GatyEtal16, IsolEtal17} may be used to refine estimates of decoded visual stimuli~\citep{StYvNase18}. Related manifold learning approaches that ``snap'' decoded estimates to the nearest surface location on a learned manifold may be used in a similar way to decode sequences of movie frames~\citep{HeusEtal18a}.  Taken together, these findings suggest that not all \textit{possible} images (and by extension, all possible stimuli) are equally likely.  Rather, the set of stimuli likely to be encountered live within subspaces or on manifolds in the relevant representational space.  The geometries of these manifolds provide insights into the statistical structure of the stimuli.  They also provide potential insights into how the brain represents those stimuli, for example to the extent that the geometries of these manifolds may be leveraged to improve brain decoding.  The shapes of these manifolds can potentially provide insights into how our thoughts are distributed over representational spaces.
